Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis

### 2023-03-14 Write-ups/Manuscript.Rmd
evaluation metrix

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

@@

As a reference method, we compare our method to Generalized Linear Mixed Models using Adaptive Gaussian Quadrature (GLMMadaptive).

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(here)
library(ROCR)

set.seed(314)
```

@@

As a reference method, we compare our method to Generalized Linear Mixed Models using Adaptive Gaussian Quadrature (GLMMadaptive). This is one of the fastest existing method developed for dynamic prediction of repeated generalized outcomes. Just like many mixed models, this method is very limited in terms of flexibility. For example, the model used for prediction of our simulated datset would simply be an linear model with one covariate indicating observation time: $g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$. While the flexibility of this mixed model can be increased using spline functions, the dimension of spline basis is also restricted by computational ability, and is unfeasible to implement under the scale of our simulated data or the complexity of our proposed method.

## Result

```{r}
load(here("Data/sim_data.RData"))
load(here("Data/SimOutput_fGFPCA.RData"))

# other quantities
mid <- unique(exp_df$bin)
N <- 500
M <- length(pred_lst)
J <- 1000

# code prediction value at observed points as NA
for(m in 1:length(pred_lst)){
  pred_lst[[m]]$pred_t200[pred_lst[[m]]$bin <= 200] <- NA
  pred_lst[[m]]$pred_t400[pred_lst[[m]]$bin <= 400] <- NA
  pred_lst[[m]]$pred_t600[pred_lst[[m]]$bin <= 600] <- NA
  pred_lst[[m]]$pred_t800[pred_lst[[m]]$bin <= 800] <- NA
}

```

```{r}
# true data
df_true <- sim_data[[1]] %>% select(id, sind_inx, eta_i)

# prediction results used for visualization
exp_df <- pred_lst[[1]]

# randomly draw four samples for visualization
rand_id <- sample(exp_df$id, 4)

exp_df %>%
  left_join(df_true %>% filter(sind_inx %in% mid) %>% rename(bin=sind_inx), by = c("id", "bin")) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_i))+
  geom_line(aes(x=bin, y = pred_t200, col = "200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t400, col = "400"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t600, col = "600"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t800, col = "800"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")

```

```{r}
runtime
# calculate ISE on binned grid
ise_mat <- matrix(NA, nrow = M, ncol = 4) # row index iteration, col index max observation time
colnames(ise_mat) <- c("t200", "t400", "t600", "t800")

for(m in 1:M){
  # true value at midpoints
  true_eta <- sim_data[[m]] %>% filter(sind_inx %in% mid) %>% select(id, sind_inx, eta_i) %>% rename(bin=sind_inx)
  pred_eta <- pred_lst[[m]] %>% select(-eta_hat)
  pred_eta <- pred_eta %>% left_join(true_eta, by = c("id", "bin"))

  ise_m <- c(
    t200 = sum((pred_eta$pred_t200-pred_eta$eta_i)^2, na.rm = T)/N,
    t400 = sum((pred_eta$pred_t400-pred_eta$eta_i)^2, na.rm = T)/N,
    t600 = sum((pred_eta$pred_t600-pred_eta$eta_i)^2, na.rm = T)/N,
    t800 = sum((pred_eta$pred_t800-pred_eta$eta_i)^2, na.rm = T)/N
  )

  ise_mat[m, ] <- ise_m
}
```
```{r}
# calculate AUC on unbinned grid
auc_mat <- matrix(NA, nrow = M, ncol = 4) # row index iteration, col index max observation time
colnames(auc_mat) <- c("t200", "t400", "t600", "t800")

for(m in 1:M){
  # true value as matrix
  Y_mat <- sim_data[[m]]  %>% select(id, sind_inx, Y) %>% rename(bin=sind_inx) %>% 
    pivot_wider(names_from = id, values_from = Y) %>%
    column_to_rownames("bin")

  # calculate AUC
  auc_mt <- rep(NA, 4) 
  for(t in c(200, 400, 600, 800)){

    # predicted value
    pred_name <- paste("pred_t", t, sep = "")
    pred_mat <- pred_lst[[m]] %>% select(id, bin, all_of(pred_name)) %>% 
      pivot_wider(names_from = id, values_from = pred_name) %>%
      column_to_rownames("bin")
    pred_full <- sapply(1:N,
                           function(x){approx(x=mid, y=pred_mat[ , x], xout = 1:J)$y})
    pred_full <- exp(pred200_full)/(1+exp(pred200_full))
    row_id <- complete.cases(pred_full)
    auc_perf <- performance(prediction(pred_full[row_id, ], Y_mat[row_id, ]), "auc")
    auc_mt[t] <- mean(unlist(auc_perf@y.values))
  }

  pred_eta <- pred_lst[[m]] %>% select(-eta_hat)
  pred_eta <- pred_eta %>% left_join(true_eta, by = c("id", "bin"))
}
```

### 2023-03-14 Write-ups/Manuscript.tex
evaluation metrix

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{Ying Jin \and Andrew Leroux}
\date{March 13, 2023}

@@

As a reference method, we compare our method to Generalized Linear Mixed
Models using Adaptive Gaussian Quadrature (GLMMadaptive).

>>>>

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{Ying Jin \and Andrew Leroux}
\date{March 14, 2023}

@@

As a reference method, we compare our method to Generalized Linear Mixed
Models using Adaptive Gaussian Quadrature (GLMMadaptive). This is one of
the fastest existing method developed for dynamic prediction of repeated
generalized outcomes. Just like many mixed models, this method is very
limited in terms of flexibility. For example, the model used for
prediction of our simulated datset would simply be an linear model with
one covariate indicating observation time:
\(g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t\). While the flexibility
of this mixed model can be increased using spline functions, the
dimension of spline basis is also restricted by computational ability,
and is unfeasible to implement under the scale of our simulated data or
the complexity of our proposed method.

\hypertarget{result}{%
\subsection{Result}\label{result}}

\includegraphics{Manuscript_files/figure-latex/unnamed-chunk-2-1.pdf}

\begin{verbatim}
## [1] 5.033985 5.134139 5.195642 5.152902 5.082466
\end{verbatim}
